---
title: "CellWalker2"
author: "Zhirui Hu"
date: "`r Sys.Date()`"
output: 
  rmarkdown::md_document:
    variant: markdown_github
vignette: >
  %\VignetteIndexEntry{CellWalker2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>"
)
```


After CellWalkR is installed, load the package:

```{r setup, message=FALSE, warning=FALSE}
library(CellWalkR)
library(data.table)
library(ggplot2)
library(Matrix)
```

```{r setseed, echo=FALSE}
set.seed(1)
```

## Use CellWalker2 for scRNA-Seq data

In this section, we show how to use CellWalker2 to annotate cells by reference cell type labels (or tree) and map similar cell type labels (or trees) using human PBMC scRNA-Seq datasets. The full data is downloaded from [CellTypist](https://celltypist.cog.sanger.ac.uk/Resources/Organ_atlas/Blood/Blood.h5ad) and we subsampled 2% cells for demonstration. To illustrate cell type annotation, we treat [Ren et al.](https://pubmed.ncbi.nlm.nih.gov/33657410/) as the reference dataset and [Yoshida et al.](https://www.nature.com/articles/s41586-021-04345-x) as unlabeled. We use cell types and marker genes obtained from Ren et al. to annotate cells in Yoshida et al.. To illustrate cell type mapping, we use CellWalker2 to identify the correspondence between cell type labels (or trees) from these two datasets.

#### load scRNA-Seq data

```{r loadData, eval=TRUE}
# 'data/SampleSingleCellRNASeq.rda' should be loaded upon loading the package
counts1 = SampleSingleCellRNASeq$counts1
counts2 = SampleSingleCellRNASeq$counts2
meta.data1 = SampleSingleCellRNASeq$meta.data1
meta.data2 = SampleSingleCellRNASeq$meta.data2


# remove cell types with too few cells (less than 5)
cts = sort(xtabs(~meta.data1$Original_annotation))
lowcts = names(cts)[cts < 5]
ind_rm = which(meta.data1$Original_annotation %in% lowcts) 
counts1 = counts1[, -ind_rm]
meta.data1 = meta.data1[-ind_rm, ]

cts = sort(xtabs(~meta.data2$Original_annotation))
lowcts = names(cts)[cts < 5]
ind_rm = which(meta.data2$Original_annotation %in% lowcts) 
counts2 = counts2[, -ind_rm]
meta.data2 = meta.data2[-ind_rm, ]
```

#### Cell type annotation

CellWalker2 use Seurat to preprocess scRNA-Seq data. User can provide marker genes for each cell type as a dataframe with columns *gene*, *cluster,* *avg_log2FC* (optional). If not provided, CellWalker2 can compute cell type marker genes from a reference scRNA-Seq dataset. Optionally, user can provide cell type tree as a phylo object. If not provided, CellWalker2 can also build a cell type tree from a reference scRNA-Seq dataset.

We compute cell type markers and the hierarchical relationship between cell type labels using reference dataset (Ren et al.). To do that, we use `Original_annotation` column in `meta.data1` as cell label. We also compute cell-to-cell similarity graph for query dataset (Yoshida et als.) as a part of the graph to run random walk. Since we only use a small subset of cells, we input cell type markers and cell type tree computed using full reference dataset (by setting `do.findMarkers = T, buildTree = T`).

```{r processRNA1, eval=TRUE}
dataset1= processRNASeq(counts1, meta.data1, group.col = 'Original_annotation', do.findMarkers = F, 
                        computeKNN = F, buildTree = T)

dataset2 = processRNASeq(counts2, do.findMarkers = F, computeKNN = T)
#markers = dataset1$markers  # if do.findMarkers = T
markers = SampleSingleCellRNASeq$RNA_markers1
tree = SampleSingleCellRNASeq$tree1
tree = ape::keep.tip(tree, unique(meta.data1$Original_annotation))# only keep the tips exist in current dataset
dataset1$tr = tree
dataset1$markers = markers
```

Then, we compute the cell-to-label edges based on standardized gene expression level of marker genes.

```{r annotateCells, eval=TRUE}
labelEdges = computeTypeEdges(dataset2$expr_norm, markers)
```

Finally, we annotate query cells by cell type labels either hierarchically organized or not. `weight1` is the edge weight ratio between cell-label edges and cell-cell edges, which controls the information that flows from labels to the cells. If it's NULL, CellWalker2 will tune this parameter by a subsample of the cells (specified by `sampleDepth`). This parameter is tuned such that similar cells will have larger influence score. The range of possible values is provided by `labelEdgeOpts`. As tuning will take longer time, we provide the option to run different value in parallel by setting `parallel = T`. If a cell type tree is provided, `wtree` specifies the edge weight between cell type labels.

with hierarchical structure:

```{r annotateCells1, eval=TRUE}
results = annotateCells(dataset2$cellGraph, labelEdges, weight1 = NULL, sampleDepth =1000,
                        labelEdgeOpts = 10^seq(-5,3,1),  tr1 = dataset1$tr,
                        wtree = 10)
```

without hierarchical structure (not run):

```{r annotateCells2, eval=FALSE}
results_notree = annotateCells(dataset2$cellGraph, labelEdges, weight1 = NULL, sampleDepth =1000,
                        labelEdgeOpts = 10^seq(-5,3,1))
```

with parallel computing in tuning weight1 (not run):

```{r annotateCells3, eval=FALSE}
results_parallel = annotateCells(dataset2$cellGraph, labelEdges, weight1 = NULL, sampleDepth =1000,
                        labelEdgeOpts = 10^seq(-5,3,1), parallel  = T,  numCores = 8)
```

We compare the cell annotation result with the orignal annotation in the query dataset:

```{r evalAnnot, eval=TRUE, fig.width=12, fig.height=9, fig.fullwidth=TRUE}
cellLabel = results[[1]]$cellLabels[rownames(meta.data2)] # cell type labels from CellWalker2
aa = data.frame('Yoshida' = meta.data2$Original_annotation, 'Ren' = cellLabel) # combine with original annotation
aa = data.table(aa)
aa = aa[, list(count= .N), by = c('Yoshida', 'Ren')]
aa[, prob:= count/sum(count), by =  'Yoshida']


aa$Ren = factor(aa$Ren, levels = SampleSingleCellRNASeq$label_ord1)
aa$Yoshida = factor(aa$Yoshida, levels = SampleSingleCellRNASeq$label_ord2)

ggplot(aa, aes(x= Ren, y=Yoshida, size=count, color=prob, group=Ren)) + 
  geom_point(alpha = 0.8) + 
  theme_bw() +theme(axis.text.x = element_text(angle = 45, hjust=1), text = element_text(size = 14)) + 
  scale_color_gradient(low = "mediumblue",  high = "red2", space = "Lab", limit = c(0, 1))+scale_size(range = c(0.5, 6)) 

```

#### Cell types (trees) mapping

To map cell types, CellWalker2 needs marker genes for each cell type and (optionally) cell type tree. We've already obtained marker genes for reference in the last section, so we only need to compute markers for query. This step can be skipped if marker genes (and tree) are provided from external data. Here, as we only input a sampled data, we input precomputed markers and cell type tree using the full data.

```{r processRNA2, eval=TRUE}
dataset2= processRNASeq(counts2, meta.data2, group.col = 'Original_annotation', do.findMarkers = F, 
                        computeKNN = F, buildTree = F) # set do.findMarkers = T, buildTree = T if compute markers and cell type tree
dataset2$markers = SampleSingleCellRNASeq$RNA_markers2

tree = SampleSingleCellRNASeq$tree2
tree = ape::keep.tip(tree, unique(meta.data2$Original_annotation))# only keep the tips exist in current dataset
dataset2$tr = tree
```

Although not the main purpose of CellWalker2, if the dataset is unlabeled, CellWalker2 can identify cell clusters first and map cell clusters among datasets.

Then, we compute the cell-to-label edges based on gene expression level of marker genes, merge the data from both datasets to compute a cell-to-cell graph. If `integrate = F`, just combine the data together without removing batch effect; if `integrate = T`, will use CCA method in Seurat to integrate data.

```{r mapCelltype, eval=TRUE}
labelEdges1 = computeTypeEdges(dataset1$expr_norm, dataset1$markers)
labelEdges2 = computeTypeEdges(dataset2$expr_norm, dataset2$markers, log2FC.cutoff = 0.25) # adjust log2FC to be smaller to increase number of markers
labelEdgesList = list(labelEdges1, labelEdges2) # make sure that cell names are different in each dataset
mergeResult =  mergeRNASeq(list(counts1, counts2), nfeatures = 5000) # make sure that cell names are different in each dataset. if there are cell names are duplicated, will add '_x' to cell names
```

Finally, we run CellWalker2 to map the cell types from two datasets either with or without the hierarchical structure of cell type labels. We set `labelEdgeWeight = NULL`, so CellWalker2 will tune the weight ratio between cell-to-label edges and cell-to-cell edges for each set of label, which controls the information that flows from labels to the cells. These parameters are tuned such that similar cells will have larger influence. By setting `parallel = T` and specifying `numCores`, CellWalker2 will tune edgeWeights in parallel. `sampleDepth` specifies the number of cells subsampled to tune these parameters. Smaller sample size will accelerate tuning but might be less optimal.

If `compute.Zscore = T`, CellWalker2 will compute Z-scores by comparing the observed influence score with its null distribution generated by permuting cell-to-label edges. Otherwise, CellWalker2 will output influence score matrix between labels only. `nround` is the number of randomization to compute Z-scores.

CellWalker2 will output influence scores or Z-scores from cell types in dataset1 to that in dataset2 and vice versa. For Z-scores, they could be different but similar.

CellWalker2 uses `%dppar%` to run each randomization parellally. `foreach` and `doParallel` packages are required and you need to register the parallel backend before running `mapCellTypes`.

```{r setup3, message=FALSE, warning=FALSE}
library(foreach)
library(doParallel)
cl<-makeCluster(8)  # change to smaller if your machine doesn't have enough cores
registerDoParallel(cl)  # for computing Z-score
```

without hierarchical structure (not run):

```{r mapCelltype2, eval=FALSE}
cellWalk2_notree = mapCellTypes(mergeResult$cellGraph, labelEdgesList, labelEdgeWeights = NULL,
                         compute.Zscore = TRUE,  nround = 55, parallel  = T,  
                         sampleDepth =1000, numCores = 8)

```

with hierarchical structure (not run):

```{r mapCelltype1, eval=FALSE}
treeList = list(dataset1$tr, dataset2$tr)
cellWalk2 = mapCellTypes(mergeResult$cellGraph, labelEdgesList, labelEdgeWeights = NULL,
                         treeList = treeList, compute.Zscore = TRUE,  nround = 55,
                         parallel  = T,  sampleDepth =1000, numCores = 8)

# if encounter error in serialize(data, node$con), can try re-initiate by:
# stopCluster(cl)
# cl<-makeCluster(8)
# registerDoParallel(cl)
```

By default, CellWalker2 will permute the edge weights between cells to each set of cell type labels independently. We can also choose which set of cell-to-label edges to permute and/or to restrict permutation within cell groups by setting `groupsList`, as sometimes we want to maintain the (phylogenetic) correlation of edge weights between cell type labels.

Here we show how to only permute edges between cells to labels of query dataset as we want to map each cell type label in query to reference and keep the hierarchical structure of labels in reference dataset. For edges between cells to labels in reference dataset, we set all the cells to group 0 so the edge weights are not permuted. For edges between cells to labels in query dataset, we set all the cells to a single group so that edge weights are permuted between all the cells to labels.

Moreover, we can adjust the edge weight going up and down the cell type tree to discourage information going across lineages. In this case, as we only consider the information flow from query to reference, for cell type tree in reference dataset, we set the edge weight going up the tree (from tips to root) to 1 and down to 0.1, and set them reversely for cell tree in the query dataset.

Though CellWalker2 will output scores in both directions, the influence scores or Z-scores from query to reference dataset should be used in this setting.

```{r mapCelltype3, eval=TRUE}
# it may take a while to run
treeList = list(dataset1$tr, dataset2$tr)
# different weights going up/down the tree
groupsList = list(rep(0, nrow(labelEdges1)), rep(1, nrow(labelEdges2)))
wtrees = matrix(c(1,0.1,0.1,1), ncol=2) # first row for dataset 1 and second for dataset 2
cellWalk2_map = mapCellTypes(mergeResult$cellGraph, labelEdgesList, labelEdgeWeights = NULL,
                         treeList = treeList, wtrees = wtrees, groupsList = groupsList,
                         compute.Zscore = TRUE, nround = 55, parallel  = T,  
                         sampleDepth =1000, numCores = 8)
```

Alternatively, we can connect all the cells to both marker sets if the genes from both datasets are mostly shared.

```{r mapCelltype-alt,eval=FALSE, class.source = 'fold-hide'}
# not run
labelEdges1 = rbind(computeTypeEdges(dataset1$expr_norm, dataset1$markers),
              computeTypeEdges(dataset2$expr_norm, dataset1$markers))
labelEdges2 = rbind(computeTypeEdges(dataset1$expr_norm, dataset2$markers),
              computeTypeEdges(dataset2$expr_norm, dataset2$markers))
labelEdgesList = list(labelEdges1, labelEdges2) # make sure that cell names are different in each dataset
```

We can check the cell type mapping and plot Z-score matrix:

```{r evalute2, eval=TRUE}
Zscore = cellWalk2_map$zscore[[2]]
res = RcppHungarian::HungarianSolver(-Zscore[1:(nrow(Zscore) -1),1:(ncol(Zscore) -1)])$pairs # no root, remove '-1' if running without tree
res[,1] = sapply(rownames(Zscore)[res[,1]], function(x) sub('_[0-9]$', '', x))
res[,2] = sapply(colnames(Zscore)[as.numeric(res[,2])], function(x) sub('_[0-9]$', '', x))
colnames(res) = c('Yoshida labels', 'Ren labels')
message("matching cell labels with Zscore (with tree): ")
print(res) # mapping cell types using Zscore
```

```{r plotMap, eval=TRUE, fig.width=12, fig.height=7, fig.fullwidth=TRUE}
bb = reshape2::melt(Zscore[1:((nrow(Zscore) + 1)/2),1:((ncol(Zscore) + 1)/2)]) # only plot tips for illustration
bb$Var1 = sub( '_[0-9]$', '',bb$Var1)
bb$Var2 = sub( '_[0-9]$', '',bb$Var2)
colnames(bb) = c('Yoshida','Ren', 'Zscores') 
bb$Ren = factor(bb$Ren, levels = SampleSingleCellRNASeq$label_ord1)
bb$Yoshida = factor(bb$Yoshida, levels = SampleSingleCellRNASeq$label_ord2)

ggplot(bb, aes(x= Ren, y=Yoshida, group=Ren)) + 
  geom_tile(aes(fill=Zscores), alpha = 0.6) + theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust=1), 
        text = element_text(size = 14)) + 
  scale_fill_gradient(low = "white",  high = "red2", space = "Lab")
```

We can also plot the Z-score on the cell type tree:

```{r plotMapTree, eval=TRUE, fig.width=12, fig.height=7, fig.fullwidth=TRUE}
tr = treeList[[1]]
tr$node.label = colnames(Zscore)[-1:-(tr$Nnode+1)]
tr$node.label = sub('_[0-9]$', '',tr$node.label)
dat = Zscore[c('NK_2', 'T CD8 EMRA_2'),, drop=F]
colnames(dat) =  sub('_[0-9]$', '',colnames(dat))
rownames(dat) =  sub('_[0-9]$', '',rownames(dat))
cl1 = ape::extract.clade(tr, 'T_CD8_c07-TYROBP:T_gdT_c14-TRDV2:3')
cl2 = ape::extract.clade(tr, 'T_CD8_c04-COTL1:T_CD4_c09-GZMK-FOS_l:3')
subtr = ape::keep.tip(tr, c(cl1$tip.label, cl2$tip.label))
subtr = ape::keep.tip(tr, c(cl1$tip.label, cl2$tip.label))
pp = CellWalkR::plotZscoreTree(subtr, dat[,c(subtr$tip.label, subtr$node.label)], cutoff = 15)
plot(pp)
```

CellWalker can also map more than two datasets (not shown here).


## Session Information

```{r session-info}
sessionInfo()
```
